Effect of Heteroscedasticity in Covariance Matrix

### 2014-12-15 thesis/draft4.tex
New additions.

@@

\section{Normal innovations \& log-normal volatilities with AR(1)
  autocorrelation}
In this subsection we consider the case
\begin{eqnarray}
  \log \sigma_t &=& \phi\log \sigma_{t-1} + x_t \nonumber \\
  &=& \sum_{l = 0}^{\infty} \phi^l x_{t-l} \label{eq:AR}
\end{eqnarray}
where $x_t \sim N(0, \theta^2)$ and we have assumed that the $\log
\sigma_t$ process was started in the infinite past with $\log
\sigma_{-\infty} = x_{-\infty}$. In addition, we assume $\log
\sigma_t$ is a stationary process with $|\phi| < 1$. Clearly $\log
\sigma_t$ is normally distributed with mean 0 and variance ${\theta^2
  / (1 - \phi^2)}$. According to \eqref{eq:greens}, the spectral
density of $\bm E$ as appears in \eqref{eq:E_def} is the same as if
$\sigma_t^2$'s were uncorrelated but $\log \sigma_t \sim N(0,
\theta^2/(1 - \phi^2))$.

To find the spectral density of $\bm C$ with normally distributed $\bm
\eta$ and uncorrelated, log-normally distributed $\sigma_t$, we use
the result of Biroli et al in \cite{biroli2007student}, which gives
the Blue function of $\bm C$ as
\begin{eqnarray*}
  B(z) &=& \frac{1}{z} + \int \frac{\sigma ^2 P(s) ds}{1-q \sigma ^2 z}
\end{eqnarray*}
where $q = N/T$ and $\sigma$ has the stationary distribution of
$\sigma_t$, whose probability density function is $P(s)$. Now we are
specifically considering lognormally distributed $\sigma_t$, so we
rewrite the above equation as
\begin{eqnarray}
B(z) &=& \int_{-\infty }^{\infty } \frac{e^{2 s} \exp \left(-\frac{s^2}{2
      v}\right)}{\sqrt{2 \pi  v} \left(1-q e^{2 s} z\right)} \,
ds+\frac{1}{z} \label{eq:LognormalBlue}
\end{eqnarray}
where $v = 1/(1 - \phi^2)$, reflecting the effect
of autocorrelation in $\log \sigma_t$'s. However, no closed form of
this integral is known to the author. To obtain the spectral density,
which is given by the Sokhotski-Weierstrass formula as
\begin{eqnarray}
  \rho(\lambda) &=& -{1 \over \pi} \lim_{\epsilon \to 0} \im G(\lambda +
  i\epsilon) \label{eq:Sokhotski-Weierstrass}
\end{eqnarray}
we equate the real part of RHS of \eqref{eq:LognormalBlue} to $\lambda \in
R$ and the imaginery part to 0, then we can solve \eqref{eq:LognormalBlue}
numerically and obtain $\rho(\lambda)$ via
\eqref{eq:Sokhotski-Weierstrass}. Specifically we solve the following
equation system:
\begin{eqnarray}
\lambda &=& \int_{-\infty}^{\infty} \frac{ds}{\sqrt{2\pi v}}
\frac{
  e^{-s^2/2v}(e^{-2s} - qx)
}{
  (e^{-2s} - qx)^2 + q^2 y^2  
} +
\frac{x}{x^2 + y^2} \label{eq:LognormalBlueReal}\\
0 &=& \int_{-\infty}^{\infty} \frac{ds}{\sqrt{2\pi v}}
\frac{
  e^{-s^2/2v} q y
}{
  (e^{-2s} - qx)^2 + q^2 y^2  
} -
\frac{y}{x^2 + y^2} \label{eq:LognormalBlueImag}
\end{eqnarray}
This is the method used by Biroli et al in
\cite{biroli2007student}. Figure \ref{fig:LognormalBlue}
shows the plot of the Blue function \eqref{eq:LognormalBlue}. $-\pi
\rho(\lambda)$ is the projection of the contour $\im B(z) = 0$ onto the $\im
G$ axis.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.45, clip=true, trim=66 0 84
    0]{../pics/LognormalBlue.eps}
  \caption{\small \it The Blue function $B$. q = 0.45, v = 3}
  \label{fig:LognormalBlue}
\end{figure}
The spectral density function $\rho(\lambda)$ obtained as numerical
solutions is shown in figure \ref{fig:LognormalSpectra} and compared
to the Marcenko-Pastur law with several values of $q = N/T$ and $v$.
\begin{figure}[htb!]
  \centering
  \subfigure[$q = 0.1$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q0.1.eps}
    \label{fig:Lognormal_q0.1}
  }
  \subfigure[$q = 0.2$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q0.2.eps}
    \label{fig:Lognormal_q0.2}
  }
  \subfigure[$q = 0.5$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q0.5.eps}
    \label{fig:Lognormal_q0.5}
  }
  \subfigure[$q = 1.0$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q1.0.eps}
    \label{fig:Lognormal_q1.0}
  }
  \caption{\small \it Spectra with lognormal volatilities. The
    empirical probability density functions are plotted as stairs. The
  density function given by the Marcenko-Pastur law is plotted as a
  thick red curve.}
  \label{fig:LognormalSpectra}
\end{figure}
From figures \ref{fig:LognormalSpectra} one can see that the
Marcenko-Pastur law is recovered when $v \to 0$, as expected.

When $v$ is small, the integrands in \eqref{eq:LognormalBlueReal} and
\eqref{eq:LognormalBlueImag} have sharp maxima at $v = 0$. Hence it is
useful to approximate them by Taylor expansion. Let's write
\begin{eqnarray*}
  \lambda &=& \int _{-\infty }^{\infty }\!g \left( s,v \right) F \left( s
  \right) {ds}+{\frac {x}{{x}^{2}+{y}^{2}}} \\
  0 &=& \int _{-\infty }^{\infty }\!g \left( s,v \right) G \left( s
  \right) {ds}-{\frac {y}{{x}^{2}+{y}^{2}}}
\end{eqnarray*}
where
\begin{eqnarray*}
g &=& \frac{1}{\sqrt{2\mathop{\rm  }\pi \mathop{\rm  }v }}e^{-\frac{s
    ^{2}}{2\mathop{\rm  }v }}\\
F \left(s \right) &=& \frac{\left(e^{-2\mathop{\rm  }s }-q \mathop{\rm
    }x \right)}{\left(e^{-2\mathop{\rm  }s }-q \mathop{\rm  }x
  \right)^{2}+q ^{2}\mathop{\rm  }y ^{2}}\mathop{\rm  } \\
G(s) &=& \frac{q \mathop{\rm  }y }{\left(e^{-2\mathop{\rm  }s }-q
    \mathop{\rm  }x \right)^{2}+q ^{2}\mathop{\rm  }y ^{2}}\\
\end{eqnarray*}

Taylor-expanding $F(s)$ and $G(s)$ to 2nd order in $s$ and plugging
them into \eqref{eq:LognormalBlueReal} and
\eqref{eq:LognormalBlueImag} gives
\begin{eqnarray}
\lambda &=& {-q x+1 \over (-q x+1)^2+q^2 y^2} + {v \over 2}\left\{
{4 \over (-q x+1)^2+q^2 y^2} -{24 (-q x+1) \over
  [(-q x+1)^2+q^2 y^2]^2} \right. \nonumber \\
&& \left. + {32 (-q x+1)^3 \over [(-q x+1)^2+q^2 y^2]^3} -8 {(-q
    x+1)^2 \over [(-q x+1)^2+q^2 y^2]^2}
\right\} \label{eq:lambda_linear}\\
0 &=& {q y \over (-q x+1)^2+q^2 y^2} + {v \over 2} \left\{
  {32 q y (-q x+1)^2 \over [(-q x+1)^2+q^2 y^2]^3}
  -{8 q y \over [(-q x+1)^2+q^2 y^2]^2} \right. \nonumber \\
  && \left.
    -{8 q y (-q x+1) \over [(-q x+1)^2+q^2 y^2]^2}
\right\} \label{eq:constraint_linear}
\end{eqnarray}
Now that we look for a solution in the limit $y \to 0$, we set $y = 0$
in \eqref{eq:lambda_linear}; in \eqref{eq:constraint_linear} we first
cancel the factor $y$ and then set $y = 0$. This yields
\begin{eqnarray}
  \lambda &=& {1 \over (-q x+1)} + {2 v (q x+1) \over (-q x+1)^3} + {1
    \over x} \label{eq:lambda_1} \\
  0 &=& {q \over (-q x+1)^2} + {4 v q(q x+2) \over (-q x+1)^4} - {1
    \over x^2} \label{eq:constraint_1}
\end{eqnarray}
Evaluating \eqref{eq:lambda_1} at $x = \lambda^{MP}_{+} = (1 +
\sqrt{q})^2$ gives a rough estimate about the expected maximum
eigenvalue, respectively. That is
\begin{eqnarray}
  \lambda_{+} &=& \lambda_{+}^{MP} [1+2v(2\sqrt q +
  1)] \label{eq:lambda_3}
\end{eqnarray}
This result is checked against simulations, as shown in figure
\ref{fig:lambda_verified}. Clearly one can see that the linear
approximation \eqref{eq:lambda_3} fits fairly well to the estimated
$\E \lambda_\M$.
\begin{figure}[htb!]
  \centering
  \subfigure[$q = 0.1$]{
    \includegraphics[scale=0.35]{../pics/lambda_max_q0.1.eps}
    \label{fig:lambda_max_q0.1}
  }
  \subfigure[$q = 0.2$]{
    \includegraphics[scale=0.35]{../pics/lambda_max.eps}
    \label{fig:lambda_max_q0.2}
  }
  \subfigure[$q = 0.5$]{
    \includegraphics[scale=0.35]{../pics/lambda_max_q0.5.eps}
    \label{fig:lambda_max_q0.5}
  }
  \subfigure[$q = 1.0$]{
    \includegraphics[scale=0.35]{../pics/lambda_max_q1.0.eps}
    \label{fig:lambda_max_q1.0}
  }
  \caption{\small \it Estimated $\E \lambda_\M$ against its linear
    approximation.}
  \label{fig:lambda_verified}
\end{figure}

>>>>

\section{Introduction}

@@

\section{Normal innovations \& log-normal volatilities with AR(1)
  autocorrelation}
In this subsection we consider the case
\begin{eqnarray}
  \log \sigma_t &=& \phi\log \sigma_{t-1} + x_t \nonumber \\
  &=& \sum_{l = 0}^{\infty} \phi^l x_{t-l} \label{eq:AR}
\end{eqnarray}
where $x_t \sim N(0, \theta^2)$ and we have assumed that the $\log
\sigma_t$ process was started in the infinite past with $\log
\sigma_{-\infty} = x_{-\infty}$. In addition, we assume $\log
\sigma_t$ is a stationary process with $|\phi| < 1$. Clearly $\log
\sigma_t$ is normally distributed with mean 0 and variance ${\theta^2
  / (1 - \phi^2)}$. The auto-correlation function $\rho(n)$ decays
exponentially: $\rho_n = \phi^{n-1} = exp[-(n-1)\ln(1/\phi)]$.
According to \eqref{eq:greens}, the spectral
density of $\bm E$ as appears in \eqref{eq:E_def} is the same as if
$\sigma_t^2$'s were uncorrelated but $\log \sigma_t \sim N(0,
\theta^2/(1 - \phi^2))$.

To find the spectral density of $\bm C$ with normally distributed $\bm
\eta$ and uncorrelated, log-normally distributed $\sigma_t$, we use
the result of Biroli et al in \cite{biroli2007student}, which gives
the Blue function of $\bm C$ as
\begin{eqnarray*}
  B(z) &=& \frac{1}{z} + \int \frac{\sigma ^2 P(s) ds}{1-q \sigma ^2 z}
\end{eqnarray*}
where $q = N/T$ and $\sigma$ has the stationary distribution of
$\sigma_t$, whose probability density function is $P(s)$. Now we are
specifically considering lognormally distributed $\sigma_t$, so we
rewrite the above equation as
\begin{eqnarray}
B(z) &=& \int_{-\infty }^{\infty } \frac{e^{2 s} \exp \left(-\frac{s^2}{2
      v}\right)}{\sqrt{2 \pi  v} \left(1-q e^{2 s} z\right)} \,
ds+\frac{1}{z} \label{eq:LognormalBlue}
\end{eqnarray}
where $v = 1/(1 - \phi^2)$, reflecting the effect
of autocorrelation in $\log \sigma_t$'s. However, no closed form of
this integral is known to the author. To obtain the spectral density,
which is given by the Sokhotski-Weierstrass formula as
\begin{eqnarray}
  \rho(\lambda) &=& -{1 \over \pi} \lim_{\epsilon \to 0} \im G(\lambda +
  i\epsilon) \label{eq:Sokhotski-Weierstrass}
\end{eqnarray}
we equate the real part of RHS of \eqref{eq:LognormalBlue} to $\lambda \in
R$ and the imaginery part to 0, then we can solve \eqref{eq:LognormalBlue}
numerically and obtain $\rho(\lambda)$ via
\eqref{eq:Sokhotski-Weierstrass}. Specifically we solve the following
equation system:
\begin{eqnarray}
\lambda &=& \int_{-\infty}^{\infty} \frac{ds}{\sqrt{2\pi v}}
\frac{
  e^{-s^2/2v}(e^{-2s} - qx)
}{
  (e^{-2s} - qx)^2 + q^2 y^2  
} +
\frac{x}{x^2 + y^2} \label{eq:LognormalBlueReal}\\
0 &=& \int_{-\infty}^{\infty} \frac{ds}{\sqrt{2\pi v}}
\frac{
  e^{-s^2/2v} q y
}{
  (e^{-2s} - qx)^2 + q^2 y^2  
} -
\frac{y}{x^2 + y^2} \label{eq:LognormalBlueImag}
\end{eqnarray}
This is the method used by Biroli et al in
\cite{biroli2007student}. Figure \ref{fig:LognormalBlue}
shows the plot of the Blue function \eqref{eq:LognormalBlue}. $-\pi
\rho(\lambda)$ is the projection of the contour $\im B(z) = 0$ onto the $\im
G$ axis.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.45, clip=true, trim=66 0 84
    0]{../pics/LognormalBlue.eps}
  \caption{\small \it The Blue function $B$. q = 0.5, v = 0.5. The
    thick black curve in the left figure is the $\im B = 0$ curve. The
    smallest (largest) eigenvalue is achieved at the left (right) end
    of the curve.}
  \label{fig:LognormalBlue}
\end{figure}
The spectral density function $\rho(\lambda)$ obtained as numerical
solutions is shown in figure \ref{fig:LognormalSpectra} and compared
to the Marcenko-Pastur law with several values of $q = N/T$ and $v$.
\begin{figure}[htb!]
  \centering
  \subfigure[$q = 0.1$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q0.1.eps}
    \label{fig:Lognormal_q0.1}
  }
  \subfigure[$q = 0.2$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q0.2.eps}
    \label{fig:Lognormal_q0.2}
  }
  \subfigure[$q = 0.5$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q0.5.eps}
    \label{fig:Lognormal_q0.5}
  }
  \subfigure[$q = 1.0$]{
    \includegraphics[scale=0.35]{../pics/Lognormal_q1.0.eps}
    \label{fig:Lognormal_q1.0}
  }
  \caption{\small \it Spectra with lognormal volatilities. The
    empirical probability density functions are plotted as stairs. The
  density function given by the Marcenko-Pastur law is plotted as a
  thick red curve.}
  \label{fig:LognormalSpectra}
\end{figure}
From figures \ref{fig:LognormalSpectra} one can see that the
Marcenko-Pastur law is recovered when $v \to 0$, as expected.

When $v$ is small, the integrands in \eqref{eq:LognormalBlueReal} and
\eqref{eq:LognormalBlueImag} have sharp maxima at $s = 0$. Hence it is
useful to approximate them by Taylor expansion. Let's write
\begin{eqnarray*}
  \lambda &=& \int _{-\infty }^{\infty }\!g \left( s,v \right) F \left( s
  \right) {ds}+{\frac {x}{{x}^{2}+{y}^{2}}} \\
  0 &=& \int _{-\infty }^{\infty }\!g \left( s,v \right) G \left( s
  \right) {ds}-{\frac {y}{{x}^{2}+{y}^{2}}}
\end{eqnarray*}
where
\begin{eqnarray*}
g &=& \frac{1}{\sqrt{2\mathop{\rm  }\pi \mathop{\rm  }v }}e^{-\frac{s
    ^{2}}{2\mathop{\rm  }v }}\\
F \left(s \right) &=& \frac{\left(e^{-2\mathop{\rm  }s }-q \mathop{\rm
    }x \right)}{\left(e^{-2\mathop{\rm  }s }-q \mathop{\rm  }x
  \right)^{2}+q ^{2}\mathop{\rm  }y ^{2}}\mathop{\rm  } \\
G(s) &=& \frac{q \mathop{\rm  }y }{\left(e^{-2\mathop{\rm  }s }-q
    \mathop{\rm  }x \right)^{2}+q ^{2}\mathop{\rm  }y ^{2}}\\
\end{eqnarray*}

Taylor-expand $F(s)$ and $G(s)$ to 2nd order in $s$
\begin{eqnarray*}
\lambda &=& F(0) + {v \over 2}F^{''}(0) + O(v^2) \\
0 &=& G(0) + {v \over 2}G^{''}(0) + O(v^2) \\
\end{eqnarray*}
Plugging them into \eqref{eq:LognormalBlueReal} and
\eqref{eq:LognormalBlueImag} gives
\begin{eqnarray}
\lambda &=& {-q x+1 \over (-q x+1)^2+q^2 y^2} + {v \over 2}\left\{
{4 \over (-q x+1)^2+q^2 y^2} -{24 (-q x+1) \over
  [(-q x+1)^2+q^2 y^2]^2} \right. \nonumber \\
&& \left. + {32 (-q x+1)^3 \over [(-q x+1)^2+q^2 y^2]^3} -8 {(-q
    x+1)^2 \over [(-q x+1)^2+q^2 y^2]^2}
\right\} \label{eq:lambda_linear}\\
0 &=& {q y \over (-q x+1)^2+q^2 y^2} + {v \over 2} \left\{
  {32 q y (-q x+1)^2 \over [(-q x+1)^2+q^2 y^2]^3}
  -{8 q y \over [(-q x+1)^2+q^2 y^2]^2} \right. \nonumber \\
  && \left.
    -{8 q y (-q x+1) \over [(-q x+1)^2+q^2 y^2]^2}
\right\} \label{eq:constraint_linear}
\end{eqnarray}
Now that we look for a solution in the limit $y \to 0$, we set $y = 0$
in \eqref{eq:lambda_linear}; in \eqref{eq:constraint_linear} we first
cancel the factor $y$ and then set $y = 0$. This yields
\begin{eqnarray}
  \lambda &=& {1 \over (-q x+1)} + {2 v (q x+1) \over (-q x+1)^3} + {1
    \over x} \label{eq:lambda_1} \\
  0 &=& {q \over (-q x+1)^2} + {4 v q(q x+2) \over (-q x+1)^4} - {1
    \over x^2} \label{eq:constraint_1}
\end{eqnarray}
Evaluating \eqref{eq:lambda_1} at $x = \lambda^{MP}_{+} = (1 +
\sqrt{q})^2$ gives a rough estimate about the expected maximum
eigenvalue, respectively. That is
\begin{eqnarray}
  \lambda_{+} &=& \lambda_{+}^{MP} [1+2v(2\sqrt q +
  1)] \label{eq:lambda_3}
\end{eqnarray}
This result is checked against simulations, as shown in figure
\ref{fig:lambda_verified}. Clearly one can see that the linear
approximation \eqref{eq:lambda_3} fits fairly well to the estimated
$\E \lambda_\M$.
\begin{figure}[htb!]
  \centering
  \subfigure[$q = 0.1$]{
    \includegraphics[scale=0.35]{../pics/lambda_max_q0.1.eps}
    \label{fig:lambda_max_q0.1}
  }
  \subfigure[$q = 0.2$]{
    \includegraphics[scale=0.35]{../pics/lambda_max.eps}
    \label{fig:lambda_max_q0.2}
  }
  \subfigure[$q = 0.5$]{
    \includegraphics[scale=0.35]{../pics/lambda_max_q0.5.eps}
    \label{fig:lambda_max_q0.5}
  }
  \subfigure[$q = 1.0$]{
    \includegraphics[scale=0.35]{../pics/lambda_max_q1.0.eps}
    \label{fig:lambda_max_q1.0}
  }
  \caption{\small \it Estimated $\E \lambda_\M$ against its linear
    approximation.}
  \label{fig:lambda_verified}
\end{figure}
